---
title: "Benchmarking LLMs for Evidence-Grounded Variable Relationship Extraction"
author:
  - name: Author Name
    affiliation:
    - &aff1 Author Institution, City, Country.
    email: email@website.com
date: "2026-01-22"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
vignette: >
  %\VignetteIndexEntry{LLM Evidence-Grounded Benchmark}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Programming environment

```{r Set random seed, include=FALSE}
seed <- 20260122
set.seed(seed)
```

```{r Load packages, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggpubr)
library(pbapply)
library(fs)
library(here)
library(readxl)
library(jsonlite)
```

```{r Project root sanity check, include=FALSE}
stopifnot(dir_exists("."))
stopifnot(dir_exists(here::here("data_registry")))
```

```{r Load functions, include=FALSE}
lapply(list.files("R/", full.names = TRUE), source)
```

```{r Set theme, include=FALSE}
my_theme_set()
```

```{r Session information, echo=FALSE}
sessionInfo()
```

# Data Analysis — Benchmark Evaluation Setup

This document initializes the data analysis environment for
downstream, read-only evaluation of adjudicated extraction outputs.
No LLM calls are performed beyond this point.

## Raw Data Inputs (Read-only)

The following data sources are treated as immutable, read-only inputs:
- Phase 1 ingestion log (`../data_registry/phase1_ingest_log.csv`)
- Phase 2 document-level logs (`../data_registry/phase2_doc_log__*.csv`)
- Phase 3 per-edge exports and judge outputs under `../data_registry/`

## Phase-3 Result Folders (Enumeration)

This subsection enumerates available Phase-3 result folders per LLM
under `../data_registry/`. No files are read or loaded at this stage.

```{r Enumerate Phase-3 folders, echo=TRUE}
phase3_dirs <- list.dirs(
  here::here("data_registry"),
  recursive = FALSE,
  full.names = TRUE
) |>
  basename() |>
  keep(~ str_detect(.x, "^phase3_"))

tibble(
  llm_phase3_folder = phase3_dirs
)
```

## Load Per-edge Adjudicated Data (Phase 3)

This subsection loads per-edge adjudicated data for each available LLM.
For each Phase-3 folder, the base export is loaded together with any
available judge outputs. No metrics are computed at this stage.

```{r Load Phase-3 per-edge data, echo=TRUE}
load_phase3_llm <- function(phase3_dir) {
  base_path <- here::here("data_registry", phase3_dir)
  
  export_file <- file.path(base_path, "export.csv")
  stopifnot(file_exists(export_file))
  
  export_df <- read_csv(export_file, show_col_types = FALSE)
  
  judge_files <- list.files(
    base_path,
    pattern = "^judge_.*\\.csv$",
    full.names = TRUE
  )
  
  judges_df <- judge_files |>
    set_names(basename) |>
    map(~ read_csv(.x, show_col_types = FALSE))
  
  list(
    llm = phase3_dir,
    export = export_df,
    judges = judges_df
  )
}

phase3_data <- phase3_dirs |>
  set_names() |>
  map(load_phase3_llm)

names(phase3_data)
```


```{r List Phase-3 files per LLM}
## Diagnostic: list files per Phase-3 folder

phase3_files <- phase3_dirs |>
  set_names() |>
  map(~ {
    fs::dir_ls(
      here::here("data_registry", .x),
      type = "file"
    ) |>
      basename()
  })

phase3_files
```


## Human Calibration Sample (300 edges)

We draw a fixed random sample of 300 extracted edges for human
calibration. Sampling is stratified by study design and LLM, and
restricted to the first 10 eligible articles per study-design stratum
(meta-analysis, randomized controlled trial, observational study) as
ordered in `data_registry/search_log.xlsx`.

The sampling chunk is intentionally set to `eval=FALSE` so it will not
run during full renders. When you want to refresh the sample, run that
chunk manually; it will overwrite the recorded sample file. All
downstream chunks only read the recorded file.

```{r Generate human 300-edge sample (manual), eval=FALSE, echo=TRUE}
## Generate and record the 300-edge human calibration sample
## (run manually; downstream chunks only read the recorded file)

set.seed(seed)

stopifnot(file_exists(here::here("data_registry", "search_log.xlsx")))

search_log_raw <- readxl::read_xlsx(
  here::here("data_registry", "search_log.xlsx")
)

names_lower <- tolower(names(search_log_raw))

pick_col <- function(pattern) {
  idx <- which(str_detect(names_lower, pattern))
  if (length(idx) < 1) {
    stop("Could not find column matching pattern: ", pattern)
  }
  names(search_log_raw)[idx[1]]
}

col_eligible <- pick_col("eligible")
col_stratum <- pick_col("strata|study[_ ]?design|design")
col_paper_id <- pick_col("paper|row|id")

search_log <- search_log_raw |>
  mutate(
    eligible_std = tolower(as.character(.data[[col_eligible]])),
    stratum_std = as.character(.data[[col_stratum]]),
    paper_id_std = as.character(.data[[col_paper_id]])
  ) |>
  filter(eligible_std %in% c("yes", "y", "true", "1")) |>
  group_by(stratum_std) |>
  slice_head(n = 10) |>
  ungroup() |>
  transmute(
    paper_id = paper_id_std,
    stratum = stratum_std
  )

extract_paper_id <- function(x) {
  if (is.na(x) || !nzchar(x)) {
    return(NA_character_)
  }
  out <- tryCatch(
    jsonlite::fromJSON(x),
    error = function(e) NA
  )
  if (is.character(out) && length(out) >= 1) {
    return(out[[1]])
  }
  NA_character_
}

phase3_exports <- phase3_data |>
  imap_dfr(~ {
    export_df <- .x$export |>
      mutate(
        llm = .x$llm,
        paper_id = map_chr(paper_id_list, extract_paper_id)
      )

    export_df
  })

eligible_edges <- phase3_exports |>
  inner_join(search_log, by = "paper_id") |>
  mutate(
    llm = as.character(llm),
    stratum = as.character(stratum)
  )

stopifnot(nrow(eligible_edges) > 0)

llm_levels <- sort(unique(eligible_edges$llm))
stratum_levels <- sort(unique(eligible_edges$stratum))

n_cells <- length(llm_levels) * length(stratum_levels)
base_n <- floor(300 / n_cells)
remainder <- 300 - base_n * n_cells

cell_plan <- tidyr::crossing(
  llm = llm_levels,
  stratum = stratum_levels
) |>
  mutate(
    n_target = base_n + if_else(row_number() <= remainder, 1L, 0L)
  )

sampled_edges <- cell_plan |>
  group_by(llm, stratum) |>
  group_modify(~ {
    cell_edges <- eligible_edges |>
      filter(
        llm == .y$llm,
        stratum == .y$stratum
      )

    n_take <- min(.x$n_target[1], nrow(cell_edges))

    if (n_take < 1) {
      return(tibble())
    }

    cell_edges |>
      slice_sample(n = n_take)
  }) |>
  ungroup() |>
  mutate(
    seed = seed,
    sampled_at_utc = format(Sys.time(), tz = "UTC", usetz = TRUE)
  )

dir_create(here::here("analysis", "derived"))
out_file <- here::here(
  "analysis",
  "derived",
  "human_calibration_sample_300_edges.csv"
)

write_csv(sampled_edges, out_file)

out_file
```

```{r Load human 300-edge sample, echo=TRUE}
sample_file <- here::here(
  "analysis",
  "derived",
  "human_calibration_sample_300_edges.csv"
)

stopifnot(file_exists(sample_file))

human_sample <- read_csv(sample_file, show_col_types = FALSE)

human_sample |>
  count(llm, stratum) |>
  arrange(llm, stratum)
```
