---
title: "Benchmarking LLMs for Evidence-Grounded Variable Relationship Extraction"
author:
  - name: Author Name
    affiliation:
    - &aff1 Author Institution, City, Country.
    email: email@website.com
date: "2026-01-22"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
vignette: >
  %\VignetteIndexEntry{LLM Evidence-Grounded Benchmark}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Programming environment

```{r Set random seed, include=FALSE}
seed <- 2026-01-22
set.seed(seed)
```

```{r Load packages, include=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggpubr)
library(pbapply)
library(fs)
library(here)
library(readxl)
library(jsonlite)
```

```{r Project root sanity check, include=FALSE}
stopifnot(dir_exists("."))
stopifnot(dir_exists(here::here("data_registry")))
```

```{r Load functions, include=FALSE}
lapply(list.files("R/", full.names = TRUE), source)
```

```{r Set theme, include=FALSE}
my_theme_set()
```

```{r Session information, echo=FALSE}
sessionInfo()
```

# Data Analysis — Benchmark Evaluation Setup

This document initializes the data analysis environment for
downstream, read-only evaluation of adjudicated extraction outputs.
No LLM calls are performed beyond this point.

## Raw Data Inputs (Read-only)

The following data sources are treated as immutable, read-only inputs:
- Phase 1 ingestion log (`../data_registry/phase1_ingest_log.csv`)
- Phase 2 document-level logs (`../data_registry/phase2_doc_log__*.csv`)
- Phase 3 per-edge exports and judge outputs under `../data_registry/`

## Phase-3 Result Folders (Enumeration)

This subsection enumerates available Phase-3 result folders per LLM
under `../data_registry/`. No files are read or loaded at this stage.

```{r Enumerate Phase-3 folders, echo=TRUE}
phase3_dirs <- list.dirs(
  here::here("data_registry"),
  recursive = FALSE,
  full.names = TRUE
) |>
  basename() |>
  keep(~ str_detect(.x, "^phase3_"))

tibble(
  llm_phase3_folder = phase3_dirs
)
```

## Load Per-edge Adjudicated Data (Phase 3)

This subsection loads per-edge adjudicated data for each available LLM.
For each Phase-3 folder, the base export is loaded together with any
available judge outputs. No metrics are computed at this stage.

```{r Load Phase-3 per-edge data, echo=TRUE}
load_phase3_llm <- function(phase3_dir) {
  base_path <- here::here("data_registry", phase3_dir)
  
  export_file <- file.path(base_path, "export.csv")
  stopifnot(file_exists(export_file))
  
  export_df <- read_csv(export_file, show_col_types = FALSE)
  
  judge_files <- list.files(
    base_path,
    pattern = "^judge_.*\\.csv$",
    full.names = TRUE
  )
  
  judges_df <- judge_files |>
    set_names(basename) |>
    map(~ read_csv(.x, show_col_types = FALSE))
  
  list(
    llm = phase3_dir,
    export = export_df,
    judges = judges_df
  )
}

phase3_data <- phase3_dirs |>
  set_names() |>
  map(load_phase3_llm)

names(phase3_data)
```


```{r List Phase-3 files per LLM}
## Diagnostic: list files per Phase-3 folder

phase3_files <- phase3_dirs |>
  set_names() |>
  map(~ {
    fs::dir_ls(
      here::here("data_registry", .x),
      type = "file"
    ) |>
      basename()
  })

phase3_files
```


## Human Calibration Sample (300 edges)

We draw a fixed random sample of 300 extracted edges for human
calibration. Sampling is stratified by study design and LLM, and
restricted to the first 10 eligible articles per study-design stratum
(meta-analysis, randomized controlled trial, observational study) as
ordered in `data_registry/search_log.xlsx`.

The sampling chunk is intentionally set to `eval=FALSE` so it will not
run during full renders. When you want to refresh the sample, run that
chunk manually; it will overwrite the recorded sample file. All
downstream chunks only read the recorded file.

```{r Generate human 300-edge sample (manual), eval=FALSE, echo=TRUE}
## Generate and record the 300-edge human calibration sample
## (run manually; downstream chunks only read the recorded file)

set.seed(seed)

stopifnot(file_exists(here::here("data_registry", "search_log.xlsx")))

search_log_raw <- readxl::read_xlsx(
  here::here("data_registry", "search_log.xlsx")
)

## search_log.xlsx columns (known):
## article_type_filter, eligible, rank, filename, ...

search_log <- search_log_raw |>
  mutate(
    paper_id = sprintf("row_%03d", row_number()),
    eligible_std = tolower(as.character(eligible)),
    stratum = as.character(article_type_filter),
    rank_num = suppressWarnings(as.numeric(rank))
  ) |>
  filter(eligible_std %in% c("yes", "y", "true", "1")) |>
  arrange(stratum, rank_num, paper_id) |>
  group_by(stratum) |>
  slice_head(n = 10) |>
  ungroup() |>
  transmute(
    paper_id,
    stratum
  )

extract_paper_id <- function(x) {
  if (is.na(x) || !nzchar(x)) {
    return(NA_character_)
  }
  out <- tryCatch(
    jsonlite::fromJSON(x),
    error = function(e) NA
  )
  if (is.character(out) && length(out) >= 1) {
    return(out[[1]])
  }
  NA_character_
}

phase3_exports <- phase3_data |>
  imap_dfr(~ {
    export_df <- .x$export |>
      mutate(
        llm = .x$llm,
        paper_id = map_chr(paper_id_list, extract_paper_id)
      )

    export_df
  })

eligible_edges <- phase3_exports |>
  inner_join(search_log, by = "paper_id") |>
  mutate(
    llm = as.character(llm),
    stratum = as.character(stratum)
  )

stopifnot(nrow(eligible_edges) > 0)

## Planned LLMs (paper list). We only run sampling after all are available.
llm_levels <- c(
  "phase3_gpt5",
  "phase3_claude45",
  "phase3_gemini3",
  "phase3_llama4",
  "phase3_mistral2",
  "phase3_deepseek3"
)

missing_llms <- setdiff(llm_levels, names(phase3_data))

if (length(missing_llms) > 0) {
  message(
    "Phase-3 exports not yet available for: ",
    paste(missing_llms, collapse = ", ")
  )
  message(
    "Proceeding with currently available Phase-3 exports. ",
    "Re-run this sampling chunk after additional Phase-3 folders appear."
  )
}

stratum_levels <- sort(unique(eligible_edges$stratum))

n_cells <- length(llm_levels) * length(stratum_levels)
base_n <- floor(300 / n_cells)
remainder <- 300 - base_n * n_cells

cell_plan <- tidyr::crossing(
  llm = llm_levels,
  stratum = stratum_levels
) |>
  mutate(
    n_target = base_n + if_else(row_number() <= remainder, 1L, 0L)
  )

## Incremental sampler (append-only):
## If a sample file already exists, keep those rows fixed and only sample
## additional edges needed for newly-available LLMs / strata.

dir_create(here::here("analysis", "derived"))
out_file <- here::here(
  "analysis",
  "derived",
  "human_calibration_sample_300_edges.csv"
)

existing_sample <- NULL
if (file_exists(out_file)) {
  existing_sample <- read_csv(out_file, show_col_types = FALSE) |>
    mutate(
      llm = as.character(llm),
      stratum = as.character(stratum)
    )
}

## Define a stable key for de-duplication across re-runs.
make_edge_key <- function(df) {
  df |>
    mutate(
      edge_key = paste0(llm, "|", doc_sha256, "|", edge_id)
    )
}

eligible_edges_keyed <- eligible_edges |>
  mutate(
    llm = as.character(llm),
    stratum = as.character(stratum)
  ) |>
  make_edge_key()

locked_keys <- character(0)
if (!is.null(existing_sample) && nrow(existing_sample) > 0) {
  existing_keyed <- existing_sample |>
    mutate(
      llm = as.character(llm),
      stratum = as.character(stratum)
    ) |>
    make_edge_key()

  locked_keys <- unique(existing_keyed$edge_key)
}

## How many are already sampled per (llm, stratum)?
already_counts <- tibble(llm = character(), stratum = character(), n = 0L)
if (!is.null(existing_sample) && nrow(existing_sample) > 0) {
  already_counts <- existing_sample |>
    count(llm, stratum, name = "n")
}

## Compute per-cell remaining targets.
cell_plan2 <- cell_plan |>
  left_join(already_counts, by = c("llm", "stratum")) |>
  mutate(
    n = replace_na(n, 0L),
    n_need = pmax(n_target - n, 0L)
  )

new_sampled_edges <- cell_plan2 |>
  filter(n_need > 0) |>
  group_by(llm, stratum) |>
  group_modify(~ {
    cell_edges <- eligible_edges_keyed |>
      filter(
        llm == .y$llm,
        stratum == .y$stratum,
        !edge_key %in% locked_keys
      )

    n_take <- min(.x$n_need[1], nrow(cell_edges))

    if (n_take < 1) {
      return(tibble())
    }

    ## group_modify() will re-attach the grouping keys (llm, stratum)
    ## so we must NOT return them from the data frame.
    cell_edges |>
      slice_sample(n = n_take) |>
      select(-llm, -stratum, -edge_key)
  }) |>
  ungroup() |>
  mutate(
    seed = seed,
    sampled_at_utc = format(Sys.time(), tz = "UTC", usetz = TRUE)
  )

sampled_edges <- bind_rows(
  existing_sample,
  new_sampled_edges
) |>
  distinct(llm, doc_sha256, edge_id, .keep_all = TRUE)

write_csv(sampled_edges, out_file)

out_file
```

```{r Load human 300-edge sample, echo=TRUE}
sample_file <- here::here(
  "analysis",
  "derived",
  "human_calibration_sample_300_edges.csv"
)

stopifnot(file_exists(sample_file))

human_sample <- read_csv(sample_file, show_col_types = FALSE)

human_sample |>
  count(llm, stratum) |>
  arrange(llm, stratum)
```

```{r Load human 300-edge sample (labeled), echo=TRUE}
sample_file_labeled <- here::here(
  "analysis",
  "derived",
  "human_calibration_sample_300_edges_labeled.csv"
)

if (file_exists(sample_file_labeled)) {
  human_sample_labeled <- read_csv(sample_file_labeled, show_col_types = FALSE)

  human_sample_labeled |>
    count(llm, stratum) |>
    arrange(llm, stratum)
} else {
  message(
    "Labeled file not found yet: ",
    sample_file_labeled,
    "\n(After you label the sample, save it to this path and re-knit.)"
  )
}
```
